% ═══════════════════════════════════════════════════════════════════════
% IntLLM Paper — MLSys 2027 / arXiv submission
%
% Per FJQ_PHASE_E_PATH_A_PAPER_OUTLINE.md v1.0 §3:
%   10-page MLSys + 11-page arXiv submission
%   Five contributions C1-C5
%   Submission target: 2026-05-25
%
% Status (2026-04-27 — Path A Week 1, sub-step 2):
%   §3 architecture     — DRAFTED THIS COMMIT
%   §4 methodology      — STUB (next sub-step)
%   §1 §2 §5-§10        — STUB (Week 1-3 calendar per outline §7)
%
% This file is the manuscript's LaTeX source. Tables are auto-generated
% from JSON artifacts via `paper/intllm/tables/*.md` + verify_intllm_tables.py;
% each numeric literal in this file MUST correspond to a registered Claim
% per the §6.9 R7 audit gate.
% ═══════════════════════════════════════════════════════════════════════

\documentclass[10pt,twocolumn]{article}

% ── Packages ──
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{listings}
\usepackage{makecell}
\usepackage{multirow}
\usepackage{balance}

% ── Code listings (consistent with v3.1 paper) ──
\lstset{basicstyle=\ttfamily\footnotesize,breaklines=true,frame=single}

% ── Theorem environments (mostly unused; kept for parity with v3.1) ──
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}

% ── Title ──
\title{IntLLM: Compiler-Verified 1.58-Bit Ternary Language Models for\\
       Operating-System Kernel Inference}
\author{Muhamad Fajar Putranto\\TaxPrime / PrimeCore.id, Jakarta, Indonesia}
\date{}

\begin{document}
\maketitle

% ═══════════════════════════════════════════════════════════════════════
% Abstract (STUB — drafted Week 3 per Path A outline §7)
% ═══════════════════════════════════════════════════════════════════════

\begin{abstract}
\textit{[Abstract pending — drafted Week 3 of Path A calendar after the
results sections (§5-§7) land. Will summarize the five contributions:
(C1) Phase D scaling chain at training-time-quant validates
Chinchilla-style behavior at 22M-74M parameter range; (C2) bilingual
ID+EN extension via Phase E1 corpus v1.0 (25.67~B tokens at 60:40);
(C3) in-kernel deployment via FajarOS Nova v3.9.0 IntLLM kernel-path;
(C4) honest negative results on outlier-handling features
(\texttt{balanced\_calib} and Hadamard rotation) at this regime, with
3-cause attribution; (C5) compiler-verified safety via
\texttt{@kernel}/\texttt{@device}/\texttt{@noinline} annotations.]}
\end{abstract}

% ═══════════════════════════════════════════════════════════════════════
\section{Introduction}
\label{sec:intro}

\textit{[STUB — drafted Week 3 of Path A calendar. Motivates the
kernel-resident LLM use case + explains why existing solutions
(BitNet~2B4T, MMfreeLM at $\geq$370M, eBPF-AI userspace offload) do
not meet the $\leq$16~MB binary + zero-heap +
deterministic-execution constraints simultaneously. Sets up C1-C5
contribution preview.]}

% ═══════════════════════════════════════════════════════════════════════
\section{Background}
\label{sec:background}

\textit{[STUB — drafted Week 3 of Path A calendar. Covers BitNet
\citep{wang2023bitnet}, BitNet~b1.58~2B4T \citep{ma2024era},
MatMul-Free LLM \citep{zhu2024matmulfree}, and the
kernel-context constraints (no heap allocation during inference,
$\leq$16~MB total binary footprint, deterministic
execution, Fajar Lang \texttt{@kernel} annotation system).]}

% ═══════════════════════════════════════════════════════════════════════
% §3 IntLLM Architecture — DRAFTED THIS COMMIT (Path A Week 1 sub-step 2)
% ═══════════════════════════════════════════════════════════════════════
\section{IntLLM Architecture}
\label{sec:arch}

IntLLM is a decoder-only language model family that extends the
MatMul-Free LLM architecture \citep{zhu2024matmulfree} with full
integer-native operation, kernel-context-friendly memory layout, and a
calibrated ternary quantization pipeline that runs end-to-end inside
the FajarOS Nova \texttt{@kernel} address space.

\subsection{Model class hierarchy}
\label{sec:arch-class}

The training-time class hierarchy is
\texttt{HGRNBitForCausalLM}~$\to$~\texttt{HGRNBitModel}~$\to$~\texttt{HGRNBitBlock},
inherited verbatim from the upstream MatMul-Free LLM repository (frozen
pin in \texttt{python/phase\_d/\_upstream/mmfreelm/}). IntLLM extends
the upstream surface in three places without modifying upstream
sources: (i)~a custom training driver
\texttt{intllm.train.train\_loop} that adds calibrated gates per scale,
checkpoint atomicity, and a step-idle watchdog (per the cross-repo
production-safety rule of Section~\ref{sec:method-trackb}); (ii)~a
quantization helper module \texttt{intllm.quant} adding RMSNorm
variants, sigmoid/SiLU lookup tables, and the per-channel adaptive
quantizer (Section~\ref{sec:arch-bitlinear}); and (iii)~an evaluation
harness \texttt{intllm.eval} aligning measurements with
\texttt{lm-evaluation-harness~v0.4.11} for baseline parity per the
research-integrity rules of Section~\ref{sec:method-integrity}.

\subsection{BitLinear forward path}
\label{sec:arch-bitlinear}

The fundamental quantized operator is
\texttt{BitLinear}~\citep{wang2023bitnet,ma2024era}, a drop-in
replacement for \texttt{nn.Linear} that quantizes both weights and
activations during training. The forward path is

\begin{equation}
\label{eq:bitlinear}
y = \text{F.linear}\bigl(\widehat{x}_{\text{quant}},\;\widehat{W}_{\text{quant}}\bigr),
\end{equation}

where the activation and weight quantizers, with straight-through
estimator (STE) gradient passthrough, are

\begin{align}
x_{\text{norm}} &= \text{RMSNorm}(x), \\
\widehat{x}_{\text{quant}} &= x_{\text{norm}} + \text{StopGrad}\bigl(q_a(x_{\text{norm}}) - x_{\text{norm}}\bigr), \\
\widehat{W}_{\text{quant}} &= W + \text{StopGrad}\bigl(q_w(W) - W\bigr).
\end{align}

The activation quantizer $q_a$ is per-token absmax 8-bit:
\begin{equation}
\label{eq:actquant}
q_a(x)_{t,c} = \frac{\bigl\lfloor x_{t,c}\cdot s_t \bigr\rceil_{[-128,127]}}{s_t},
\quad
s_t = \frac{127}{\max_{c'} |x_{t,c'}| + \epsilon},
\end{equation}
where $\lfloor\cdot\rceil_{[-128,127]}$ denotes round-then-clamp.
The weight quantizer $q_w$ is per-tensor absmean ternary
$\{-1, 0, +1\}$ per b1.58:
\begin{equation}
\label{eq:wquant}
q_w(W)_{i,j} = \frac{\bigl\lfloor W_{i,j}\cdot s_w \bigr\rceil_{[-1,1]}}{s_w},
\quad
s_w = \frac{1}{\overline{|W|} + \epsilon},
\end{equation}
giving an effective $\log_2 3 \approx 1.58$ bits per weight after entropy
coding. We use $\epsilon = 10^{-5}$ throughout, matching the upstream
implementation.

\subsection{Per-block layout}
\label{sec:arch-block}

Each \texttt{HGRNBitBlock} contains exactly six BitLinear instances:
four in the attention sub-layer (\texttt{i\_proj}, \texttt{f\_proj},
\texttt{g\_proj}, \texttt{o\_proj}) implementing the gated linear
recurrence of MMfreeLM~\citep{zhu2024matmulfree}, and two in the
feed-forward sub-layer (\texttt{mlp.gate\_proj},
\texttt{mlp.down\_proj}). A top-level \texttt{lm\_head} BitLinear
projects to the vocabulary, giving a per-model BitLinear count of
$6L + 1$, where $L$ is the number of layers. For our Mini configuration
($L = 6$), this is $37$ BitLinear sites; this count is recoverable as a
runtime invariant via \texttt{intllm.qat.is\_bitlinear}-based traversal
and is reported by every Phase~D ablation harness.

The structural distinction between \texttt{i\_proj}/\texttt{f\_proj}/\texttt{g\_proj}
and \texttt{o\_proj} is load-bearing for our negative-result analysis
in Section~\ref{sec:negatives}: the first three implement gated linear
recurrence projections (structurally distinct from the QKV projections
of softmax attention) while \texttt{o\_proj} is the post-recurrence
output projection. Outlier-handling rotations from the QuaRot family
\citep{ashkboos2024quarot,liu2024spinquant} are canonically applied
to QKV inputs in transformers; we test the closest analogue (\texttt{o\_proj}
input rotation) at \mbox{Mini scale} and find it insufficient
(Section~\ref{sec:negatives}).

\subsection{Tokenizer and vocabulary}
\label{sec:arch-tokenizer}

We use the Mistral~v3 byte-pair encoding tokenizer with 32{,}768 token
vocabulary, chosen because (i)~it has wide multilingual coverage at a
size that fits the kernel-binary budget, (ii)~it is the same tokenizer
used by \citet{zhu2024matmulfree} so cross-paper benchmarks are
directly comparable, and (iii)~its training data spans Latin-script
languages including Indonesian, supporting the bilingual extension of
Section~\ref{sec:bilingual} without retraining the tokenizer.

\subsection{Configuration matrix}
\label{sec:arch-config}

Three IntLLM scales validate the Phase~D scaling-chain
contribution~(C1, Section~\ref{sec:results-scaling}). All three share
the BitLinear forward path of Section~\ref{sec:arch-bitlinear} and the
per-block layout of Section~\ref{sec:arch-block}; they differ only in
hidden dimension and depth.

\begin{table}[t]
\centering
\small
\caption{IntLLM configuration matrix. Parameter counts include the
ternary BitLinear weights, FP32 norms, and the embedding+lm\_head
matrices at full precision in the trained checkpoint. The
\texttt{.fjm}~v9 export format reduces the trained checkpoint to a
ternary-packed kernel binary.}
\label{tab:config}
\begin{tabular}{lrrrr}
\toprule
Config  & $d$  & $L$ & Vocab  & Params \\
\midrule
Mini    & 256  & 6   & 32{,}768 & 21.9~M \\
Base    & 384  & 12  & 32{,}768 & 46.4~M \\
Medium  & 512  & 12  & 32{,}768 & 74.5~M \\
\bottomrule
\end{tabular}
\end{table}

The hidden dimensions $\{256, 384, 512\}$ are deliberately chosen as
powers-of-two-friendly values: $d = 2^k$ supports the Walsh-Hadamard
rotation construction of Section~\ref{sec:negatives} without padding,
$d \cdot 4 / 3$ MLP intermediate sizes pack cleanly into ternary kernel
buffers, and the embedding tables fit in $\leq 2^{15}$ vocabulary
without dynamic sizing.

\subsection{Integer-native activation pipeline}
\label{sec:arch-pipeline}

Beyond BitLinear, three additional surface-area changes make the
forward path integer-native end-to-end:

\paragraph{$\gamma$-less RMSNorm.} The upstream MatMul-Free RMSNorm
multiplies its normalized output by a learnable $\gamma$ tensor; we
absorb this scale into the downstream BitLinear's effective $\beta$
calibration constant during checkpoint export, eliminating a per-block
FP multiplication from the kernel-resident forward path. No
training-time loss-of-expressivity penalty was observed in the Phase~D
scaling chain (Section~\ref{sec:results-scaling}).

\paragraph{Sigmoid and SiLU lookup tables.} The HGRN gating activation
$\sigma$ and the MLP gate activation SiLU are replaced at inference
time by 257-entry linear-interpolation lookup tables defined over
$x \in [-8, +8]$, with maximum absolute error $\leq 2 \cdot 10^{-3}$ and
$\leq 1.6 \cdot 10^{-2}$ respectively. The training-time forward path
uses standard PyTorch \texttt{torch.sigmoid} and
\texttt{torch.nn.functional.silu} for gradient stability; the
training-versus-inference equivalence is verified by an
\texttt{fp16-vs-ternary parity gate} (deferred to Path~A Week~3
calendar; Section~\ref{sec:method-trackb}).

\paragraph{Activation quantization at module boundary.} The activation
quantizer $q_a$ of Equation~\ref{eq:actquant} is applied at every
BitLinear input, producing 8-bit integer activations that are
multiplied by ternary weights and accumulated into 16-bit (training) or
32-bit (inference) integers before dequantization. At no point in the
forward path does the kernel-resident inference code path require a
floating-point matmul, satisfying the kernel-context invariant of
zero-FP-hot-path execution.

\subsection{Comparison with prior architectures}
\label{sec:arch-comparison}

Table~\ref{tab:arch-comparison} positions IntLLM against the closest
prior architectures along the axes that drive the kernel-deployment
constraint (full citations in
\texttt{docs/INTLLM\_RELATED\_WORK.md}~\textsection7).

\begin{table}[t]
\centering
\small
\caption{Architecture comparison on axes relevant to kernel-context
deployment. ``$\checkmark$'' = present, ``$\circ$'' = partial,
``$-$'' = absent. The integer-native column requires
zero floating-point matmuls in the forward path.}
\label{tab:arch-comparison}
\begin{tabular}{l c c c c c}
\toprule
                  & Tern.\ wts & Int.\ acts & No softmax & No FP $\sigma$ & Kernel \\
\midrule
BitNet b1.58 2B4T &$\checkmark$&     $-$    &     $-$    &       $-$      &  $-$  \\
MatMul-Free 2.7B  &$\checkmark$&     $-$    & $\checkmark$&      $-$      &  $-$  \\
TerEffic FPGA     &$\checkmark$&  $\circ$   &     $-$    &       $-$      &  $-$  \\
\textbf{IntLLM}   &$\checkmark$&$\checkmark$&$\checkmark$ &$\checkmark$ &$\checkmark$\\
\bottomrule
\end{tabular}
\end{table}

The composition of the five $\checkmark$'s in IntLLM's row, rather than
any single axis, is the architectural contribution. We do not claim
that any axis is novel in isolation; we claim that the joint
satisfaction of all five is what unlocks the kernel-resident deployment
demonstrated in Section~\ref{sec:results-kernel}.

% ═══════════════════════════════════════════════════════════════════════
% §4 Training Methodology and Track B — DRAFTED THIS COMMIT
% (Path A Week 1 sub-step 3)
% ═══════════════════════════════════════════════════════════════════════
\section{Training Methodology and Track~B Interruption Safety}
\label{sec:method}

The IntLLM training pipeline is structured around five methodological
commitments that together support the C1~scaling-chain and
C4~negative-results contributions: calibrated per-scale val\_loss gates
(\textsection\ref{sec:method-gates}), a bilingual training stream
matching the Q5~baseline protocol (\textsection\ref{sec:method-stream}),
a six-layer interruption-safety harness
(\textsection\ref{sec:method-trackb}), a like-for-like 24K-step ablation
budget (\textsection\ref{sec:method-budget}), and a set of
research-integrity rules adopted from the project's
\texttt{CLAUDE.md~\textsection6.9}~ruleset
(\textsection\ref{sec:method-integrity}).

\subsection{Calibrated per-scale val\_loss gates}
\label{sec:method-gates}

Each IntLLM configuration ships with a single mechanical pass/fail
threshold on held-out validation loss, against which the trained
checkpoint is judged. The thresholds are
\textbf{Mini~$<$~4.5}, \textbf{Base~$<$~4.2}, and
\textbf{Medium~$<$~4.0} for the SlimPajama-derived English validation
stream (50~batches at \texttt{seed=999}). The thresholds are
\emph{calibrated} against empirical Phase~D evidence rather than
chosen \emph{a priori}: an earlier Mini~v1 run used the uncalibrated
target $<$~4.0 and FAILED ($\text{val\_loss} = 4.46$); the same
checkpoint, re-evaluated against the calibrated Mini target $<$~4.5,
PASSES with margin~$0.04$~nat. Mini~v2 widens that margin to
$0.12$~nat, Base~c.1 to $0.21$~nat, and Medium~c.1 to $0.28$~nat. The
\emph{monotonic widening} of margins across scales is the
scaling-chain claim of contribution~C1
(Section~\ref{sec:results-scaling}); we treat it as the principal
structural signal in the paper rather than any single absolute
val\_loss number, because absolute losses are sensitive to tokenizer
choice and validation-stream sampling that other ternary papers do not
hold fixed.

Each gate decision is committed as a separate decision document
(\texttt{FJQ\_PHASE\_D\_MINI\_GATE.md},
\texttt{\_BASE\_GATE.md}, and \texttt{\_MEDIUM\_GATE.md}) per the
``decisions are mechanical, not paragraphs'' rule of
Section~\ref{sec:method-integrity}, R6.

\subsection{Bilingual training stream}
\label{sec:method-stream}

For the bilingual extension (contribution~C2,
Section~\ref{sec:bilingual}), the Phase~E1 corpus
v1.0~(25.67~B~tokens, $60$:$40$~ID:EN, 0\%~synthetic, 0.0254\%
exact-hash dedup) feeds an interleaver
\texttt{intllm.data.bilingual\_stream(id\_share=0.6)} that mixes the
Indonesian parquet shards and a streaming SlimPajama English source at
per-batch granularity via a seeded coin flip. Distinct random seeds
are used for the Indonesian source (\texttt{seed}), English source
(\texttt{seed+1}), and the per-batch language-selection RNG
(\texttt{seed+2}) so that the coin-flip cannot accidentally correlate
with shard ordering on either side.

The \emph{Q5 baseline protocol} that all subsequent E2 ablations
compare against is a Mini-scale training run on this bilingual stream
for $24{,}000$~steps at batch size~$8$ and sequence length~$1024$,
yielding $\approx196$~M tokens of training data at the design 60:40
ratio. The protocol is implemented in
\texttt{python/phase\_d/scripts/q5\_bilingual\_baseline.py} and
produces the artifact
\texttt{paper/intllm/ablations/q5\_bilingual\_baseline.json}, against
which all E2.x ablation rows are measured.

\subsection{Track~B six-layer interruption safety}
\label{sec:method-trackb}

A Phase~D Base~c.1 training run on
2026-04-22 hung for 8.5~hours after a laptop suspension event killed
the HuggingFace CDN socket beneath the streaming dataloader. The
training loop continued executing but produced zero step progress; no
checkpoint had been saved and no watchdog detected the hang. The
incident motivated a six-layer interruption-safety rule that we
codify here for any training script in the IntLLM~/ FajarQuant tree:

\begin{enumerate}[leftmargin=*]
\item \textbf{Periodic atomic checkpointing.} Every $N$~steps
($N=10{,}000$ for the Phase~D Mini/Base/Medium runs), the model,
optimizer, and LR-scheduler states are serialized via
\texttt{torch.save} to \texttt{ckpt\_step\_\{step:06d\}.pt.tmp},
then renamed via \texttt{os.replace}, which is POSIX-atomic on the
same filesystem. The next-write-wins property of \texttt{os.replace}
guarantees no half-written checkpoint is observed by a resuming
reader, even under SIGKILL during the write.

\item \textbf{Bit-exact resume.} A \texttt{--resume <path>} CLI flag
loads the model state-dict, optimizer state, and LR-scheduler state
from a checkpoint and continues training with the step counter
restored. The unit test \texttt{test\_train\_watchdog.py::}
\texttt{test\_resume\_bit\_exact\_loss} verifies that pre-save and
post-resume one-step losses match within $10^{-4}$.

\item \textbf{Step-idle watchdog.} A daemon thread polls the step
counter every 30~seconds; if the counter has not advanced for
1{,}800~seconds (30~minutes), it sends \texttt{SIGTERM} to the main
process. Combined with layer~1 and layer~2 above, the operator can
restart the training script with \texttt{--resume-auto} and lose at
most $\sim$1.5~$\times$ \texttt{ckpt\_every} steps of compute per
incident.

\item \textbf{Per-chunk read timeouts on streaming sources.} The
environment variables
\texttt{HF\_DATASETS\_DOWNLOAD\_TIMEOUT=60} and
\texttt{HF\_HUB\_DOWNLOAD\_TIMEOUT=60} are set on import of
\texttt{intllm.data}, capping any single HTTP chunk read at 60~s; a
dead socket raises a timeout instead of hanging indefinitely.

\item \textbf{Streaming retry on transient network errors.} A wrapper
\texttt{intllm.data.\_retry\_iter} reconstructs the streaming
iterator on transient exceptions (\texttt{ConnectionError},
\texttt{OSError}, \texttt{requests.Timeout},
\texttt{urllib3.ReadTimeoutError}, et al.), with exponential backoff
capped at 60~s and a per-retry seed offset so retries see a
different shuffle order. Non-retryable exceptions (\texttt{ValueError},
\texttt{KeyError}) propagate immediately.

\item \textbf{Regression gate.} A Makefile target
\texttt{make test-train-watchdog} runs $\sim$24~unit tests covering
the watchdog real-thread fire path, default \texttt{SIGTERM}
delivery, checkpoint rotation, \texttt{--resume} bit-exact load, and
\texttt{retry\_iter} on a synthetic transient exception. A
pre-push hook gates training-code changes on this target's pass.
\end{enumerate}

The harness was validated end-to-end during a manual laptop-shutdown
event on 2026-04-23 mid-Medium-training: the watchdog fired, the
checkpoint was preserved, \texttt{--resume-auto} bit-exact-restored
the model, and only $\sim$36~minutes of compute was lost out of a
$\sim$10~hour run. Subsequent Phase~E2 ablations
(Sections~\ref{sec:bilingual} and~\ref{sec:negatives}) used the same
harness without modification.

\subsection{Like-for-like ablation budget}
\label{sec:method-budget}

Every E2.x ablation in the present paper trains for exactly the same
24{,}000-step budget at the same batch size and sequence length as the
Q5 baseline of Section~\ref{sec:method-stream}, i.e.\ a constant
$\approx$196~M-token training-token budget. The budget is exposed as
the constant \texttt{E2\_DEFAULT\_N\_STEPS=24\_000} at the top of the
ablation harness driver \texttt{train\_mini\_ablation.py}; an earlier
draft of the driver defaulted to the larger 60{,}000-step Phase~D~Mini
target, which was caught and fixed during the Hadamard ablation launch
(commit \texttt{ed5a7c2}) when it was noticed that an EN-only
60{,}000-step run was not directly comparable to the bilingual
24{,}000-step Q5 baseline.

The orthogonal flag \texttt{--bilingual-data} switches the training
stream from EN-only SlimPajama to the bilingual mix described in
Section~\ref{sec:method-stream}, so any E2 feature flag
(\texttt{--hadamard}, \texttt{--balanced-calib}, etc.) can be ablated
against either the EN-only Phase~D~Mini-gate baseline or the bilingual
Q5 baseline. The negative-results experiments of
Section~\ref{sec:negatives} use \texttt{--bilingual-data} throughout
because that is the regime the paper's bilingual contribution lives in.

The principal val\_loss comparison metric across all E2 rows is
\texttt{val\_loss(EN, seed=999, 50~batches)}, identical to the Q5
baseline's English measurement, so each ablation reports a single
delta against a single baseline rather than a vector of weakly
comparable numbers.

\subsection{Research-integrity rules adopted from CLAUDE.md \textsection6.9}
\label{sec:method-integrity}

The methodology section closes with the seven research-integrity
rules adopted verbatim from the project's
\texttt{CLAUDE.md~\textsection6.9} ruleset, since those rules
materially constrain how each contribution in this paper was produced:

\begin{description}[leftmargin=*]
\item[R1 Canonical-protocol benchmarks.] All comparison benchmarks
use \texttt{lm-evaluation-harness~v0.4.11} with no custom protocol
deviation. The harness SHA is pinned in
\texttt{eval/HARNESS\_SHA} and CI fails on drift.

\item[R2 Literature review precedes algorithm design.] The
related-work synthesis (\texttt{docs/INTLLM\_RELATED\_WORK.md},
18~papers across 7~clusters) was committed before the results
sections of this paper were drafted, satisfying the
``literature first'' invariant.

\item[R3 Baseline parity (port full features, no strawman).]
Comparison baselines (BitNet~b1.58~2B4T, MMfreeLM~370M/1.3B/2.7B,
SmolLM2~135M/360M, Pythia~160M) are evaluated as published; any
unported feature is recorded in
\texttt{docs/BASELINE\_UNPORTED\_FEATURES.md} so reviewers can adjust
their judgments rather than be misled by silent strawman baselines.

\item[R4 Calibrated~$>$~per-chunk for data-driven decompositions.]
The BitLinear weight-quant scale~$s_w$ is calibrated once at QAT
freeze rather than recomputed per forward pass, eliminating a
correlated noise source on the activation-quantization grid.

\item[R5 Outlier-handling required for LLM quantization.] Every
ablation in Section~\ref{sec:negatives} has an explicit outlier-
handling rationale and adoption gate; the negative-result attribution
of Section~\ref{sec:negatives} is itself an instance of this rule's
mechanical-gate enforcement.

\item[R6 Results section last.] The architecture
(Section~\ref{sec:arch}) and methodology (this section) were drafted
before any results table was populated; results sections in this
paper draw exclusively from the JSON artifact directory
\texttt{paper/intllm/results/} and are auto-generated rather than
hand-edited.

\item[R7 Pre-publication audit gate.] The script
\texttt{scripts/verify\_intllm\_tables.py~--strict} is registered as
a pre-push hook and a CI required check; every floating-point
literal in this manuscript's tables maps to a registered claim
backed by a JSON cell, with the script failing on drift between
table and source.
\end{description}

The combination of mechanical gates (R6, R7) and committed-decision
artifacts (R5 + the per-scale gate decision documents of
Section~\ref{sec:method-gates}) is what gives the paper's negative
results their specific weight: the failures of
Section~\ref{sec:negatives} cannot be explained away as ``we tried
something and it didn't work,'' because the gates that ruled them out
were specified \emph{before} the experiments ran and ran
\emph{mechanically} on the artifacts rather than through interpretive
re-reading after the fact.

% ═══════════════════════════════════════════════════════════════════════
\section{Phase~D Scaling-Chain Results (C1)}
\label{sec:results-scaling}

\textit{[STUB — drafted Week 2. Tables~1+2, Figure~1, narrative.]}

% ═══════════════════════════════════════════════════════════════════════
\section{Bilingual Extension and Q5 Baseline (C2)}
\label{sec:bilingual}

\textit{[STUB — drafted Week 2. Phase~E1 corpus v1.0 description,
Q5 measurement protocol, Table~5 partial, Figure~4.]}

% ═══════════════════════════════════════════════════════════════════════
\section{Honest Negative Results on Outlier-Handling Features (C4)}
\label{sec:negatives}

\textit{[STUB — drafted Week 2. Lifts content from
\texttt{FJQ\_PHASE\_E\_E2\_BILINGUAL\_CALIB\_DECISION.md} v1.0 and
\texttt{FJQ\_PHASE\_E\_E2\_HADAMARD\_DECISION.md} v1.0, including the
3-cause attributions and Phase F.5/F.6 future-work pointers. Tables~5
full and~6, Figure~3.]}

% ═══════════════════════════════════════════════════════════════════════
\section{In-Kernel Deployment via FajarOS Nova (C3)}
\label{sec:results-kernel}

\textit{[STUB — drafted Week 3. FajarOS Nova v3.9.0 IntLLM kernel-path;
Table~4 binary-size + tok/s + watts; Figure~2 binary budget breakdown;
4-invariant regression gate \texttt{make test-intllm-kernel-path}.]}

% ═══════════════════════════════════════════════════════════════════════
\section{Compiler-Verified Safety (C5)}
\label{sec:safety}

\textit{[STUB — drafted Week 3. \texttt{@kernel} / \texttt{@device} /
\texttt{@noinline} annotations; type-system enforcement of the
no-heap, no-tensor, no-raw-pointer invariants at the kernel-LLM
deployment boundary; Table~7 compile-time error examples (KE001
HeapAllocInKernel, KE002 TensorInKernel, etc.); Figure~5 compilation
pipeline diagram with the @kernel analyzer pass.]}

% ═══════════════════════════════════════════════════════════════════════
\section{Discussion and Future Work}
\label{sec:discussion}

\textit{[STUB — drafted Week 3. Phase F.5 PTQ-style post-training
calibration; Phase F.6 post-hoc QuaRot on pre-trained checkpoint;
Phase F.7-F.9 deferred E2.2-E2.5 sub-features; Phase~E3 bilingual
scale-up to Base/Medium; Phase~F~Tier-3 tax-vertical; Phase~D Stretch
training. Each Phase~F entry maps to a sub-task table in
\texttt{FJQ\_PHASE\_F\_TAX\_VERTICAL\_ROADMAP.md} v1.3.]}

% ═══════════════════════════════════════════════════════════════════════
% Bibliography (BibTeX entries to be generated from
% docs/INTLLM_RELATED_WORK.md §10 References + verified against arXiv
% per CLAUDE §6.9 R7 audit gate at first-draft time)
% ═══════════════════════════════════════════════════════════════════════

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
